The Astropy Project: Building an open-science project and status of the v2.0 core package

### 2018-03-02 response.txt

> The phase 'open development' is used a few times but isn't really defined or
> connected to the broader open source ecosystem.  What other communities are
> working in this way? How novel is 'open development' by their assessment? For
> example, other communities have adopted policies that encourage community
> contribution - most notable is the Node.js Foundation's contribution policies
> designed to promote 'Healthy Open Source'
> https://medium.com/the-node-js-collection/healthy-open-source-967fa8be7951 .
> How does the Astropy community efforts compare to this?

> A number of times I found myself wondering how the Astropy projects decides
> whether to build their own package or leverage something available in the
> scientific Python ecosystem? On the face of it it seems like every decision
> made by the Astropy project to build a new core package only increases the
> long term cost/overhead of maintaining this functionality. For example,
> Section 3.5.2 describes `astropy.table` and notes that there's similarities
> to the pandas package and numpy arrays. The functionality described in
> `astropy.nddata` also sounds somewhat generic. Some discussion as to why
> existing community tools (such as pandas) weren't used would be very welcome.
> In addition, if the answer to this question is that these other tools don't
> support all of the functionality the Astropy community needs (e.g. ability to
> handle Quantities with units in tables) are any efforts being made by the
> Astropy community to contribute these changes to packages such as Pandas?

This is an important point, but turns out to be very difficult to generalize how these decisions are made (ironically!). From a *dependency* point of view, Astropy deeply intertwines with the existing Python scientific ecosystem. Where possible we always try to use other tools instead of building our own. Some major astropy functionality is essentially an astro-relevant "wrapper" around existing code. For example, modeling makes it easier to use fitting functions from scipy with models that are primarily of interest in astronomy. Other bits of functionality are really astro-specific or the work to make it more "general" is difficult.  The coordinates and nddata subpackages are good examples of this latter case - despite the general-sounding names, they are focused on the needs of the astronomy community.  The `nddata` case the referee mentions is particularly notable because the actual underlying machinery is built completely on `numpy`, but essentially provides a standard "metadata" language for N-dimensional data *in astronomy*.  That is, it has attributes like `wcs`, `flags`, etc, that have a particular meaning and shared "cultural" understanding in astronomy that is lost if it is made into a more "general" tool.

Other pieces were originally more astro-specific or had no equivalent when they were started and have grown more generalizable as time goes on - e.g. units and table. In general, it's considered "best effort" to generalize these things because the first priority is basic tools for astronomy. Such changes come with non-trivial added cost because that means it now needs a maintainer to do releases, review pull requests, maintain the docs, etc. Using the astropy community helps this, of course, but it is still more work than simply leaving these things in the core. So how much anything is generalized dependes heavily on the time/interest of an individual to do the porting. So for example some of the documentation and testing machinery is now (after v2.0) getting moved from astropy into more generally-useful independent packages.  Meanwhile, units/Quantity would require a much larger effort to disentangle from astropy and maintain independently - but no maintainer with enough time has thus far volunteered.

We have added a new section 2.7 outlining some of the above discussion.

Additionally, as the referee points out, `astropy.table` represents a particularly unusual case for this.  It falls under the "historical" case in that it was started before Pandas.  By the time Pandas became a standard tool in the pydata stack, tables had already developed a fairly large user base in astronomy, so it was not an option to completely eliminate it and say "use Pandas".  It had also developed several features missing from Pandas but important for astronomy users (e.g., support for units, fully-fledged objects as columns, and support for the FITS table concept of "multidimensional columns").  The `astropy.table` investigated this possibility, but it was clear that some of the features would require major incompatible changes *in Pandas* to do this, and it was clear by then that the Pandas user base was too large to accept this.  This led to the current solution of `to_pandas` and `from_pandas`, which allows transformation back-and-forth for the use cases where this is supported. We have added a briefer summary of this discussion to the text in the tables subsection (3.5.2).

### 2018-03-02 main.tex
Typo fix from @crwafordsm's feedback

The \astropy project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy project.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g.,
\href{http://docs.astropy.org/en/stable/nddata/index.html}{\texttt{NDData}}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

In order to check the accuracy of the coordinate transformations in
\astropysubpkg{coordinates}, we have created a set of benchmarks that we use
to compare transformations between a set of coordinate frames for a number of
packages\footnote{\url{http://www.astropy.org/coordinates-benchmark/summary.html}}.
Since no package can be guaranteed to implement all transformations to
arbitrary precision and some transformations are sometimes subject to
interpretation of standards (in particular in the case of Galactic coordinates),
we do not designate any of the existing packages as the ``ground truth'' but
instead compare each tool to all other tools. The benchmarks are thus useful
beyond the \astropy project since they allow all of the tools to be compared to
all other tools. The tools included in the benchmark at the moment include the
\astropypkg core package, Kapteyn \citep{kapteyn}, NOVAS \citep{novas},
PALpy \citep{pal}, PyAST \citep[a wrapper for AST, described in][]{ast},
PyTPM\footnote{\url{https://github.com/phn/pytpm}}, PyEphem \citep{pyephem},
and pySLALIB \citep[a Python wrapper for SLALIB, described in][]{slalib}.

@@

The \astropysubpkg{table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} dataframes \citep{pandas} but with modifications for
astronomical data. Most notably, tables from \astropysubpkg{table}
allow for table or column metadata and can handle
vectors or arrays as table entries.
The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \astropysubpkg{table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.

>>>>

The \astropy Project, as of the version 2.0 release, does not receive any direct
financial support for the development of \astropypkg.
Development of the software, all supporting materials, and community support are
provided by individuals who work on the \astropy Project in their own personal
time, by individuals or groups contributing to \astropy as part of a research
project, or contributions from institutions that allocate people to work on
\astropy.
A list of organizations that have contributed to \astropy in this manner
can be found in the Acknowledgments.

Different funding models have been proposed for support of \astropy
(e.g., \citealt{2016arXiv161003159M}), but a long-term plan
for sustainability has not yet been established.
The \astropy Project has the ability to accept financial contributions
from institutions or individuals through the NumFOCUS\footnote{NumFOCUS
is a 501(c)(3) nonprofit that supports and promotes world-class, innovative,
open source scientific computing.} organization. NumFOCUS has, to date, covered the direct costs incurred by the \astropy Project.

\subsection{Reuse of the scientific Python ecosystem}

The \astropy Project is built on a philosophy of building from the existing
Python scientific ecosystem wherever possible. Many of the enabling technologies
like \texttt{numpy}, \texttt{scipy}, \texttt{matplotlib}, or \texttt{cython},
are necessary as the core underpinnings of \astropy packages.  Often this means
using these packages as ``building blocks'' (e.g., the ubiquitous
use of \texttt{numpy} arrays throughout \astropypkg).  In other cases this means
wrappers around more general algorithms that make them more convenient for
astronomy use cases (e.g., the model-fitting in the \package{astropy.modeling}
subpackage discussed in \sectionname~\ref{sec:modeling}).  Sometimes these
simple wrappers evolve into more complex implementations that address
astronomically-relevant use cases the general tool does not support (e.g.,
\package{astropy.convolution}, see \sectionname~\ref{ssec:convolution}). As a
broad rule, the Project explicitly encourages re-use of code where possible.

However, the boundaries of when this re-use is called for is often ambiguous.  Some
of the examples above only evolved after significant debate in the community
over whether these algorithms were sufficient.  Other times, even apparently
general functionality did not exist in the wider ecosystem that met the \astropy
community's needs, and hence the functionality had to be developed wholly from
the developer resources available in the community (e.g.,
\package{astropy.units}, or \package{astropy.table} when it began). At the same 
time, however, the \astropy Project has provided the wider community with myriad bug
fixes to the enabling technologies listed above, as well as the testing and
documentation architecture. Functionality is only ``extracted'' from \astropy to 
other packages after careful consideration that includes considering the impact 
on the maintenance and support of \astropy.

\section{Astropy Core Package Version 2.0}
\label{sec:core}
The \astropy Project aims to provide \python-based packages for all tasks that
are commonly needed in a large subset of the astronomical community.
At the foundation is the \astropypkg core package, which provides general
functionality (e.g., coordinate transformations, reading and writing
astronomical files, and units) or base classes for other
packages to utilize for a common interface (e.g.,
\href{http://docs.astropy.org/en/stable/nddata/index.html}{\texttt{NDData}}).
In this section, we highlight new features introduced or substantially improved
since version 0.2 (previously described in \citealt{astropy}).  The \astropypkg
package
provides a full log of changes\footnote{\url{https://github.com/astropy/astropy/blob/stable/CHANGES.rst}}
over the course of the entire project and more details about individual
subpackages are available in the documentation.\footnote{\url{http://docs.astropy.org/en/stable/}}
Beyond what is mentioned below, most subpackages have seen improved performance
since the release of the version 0.2 package.

@@

In order to check the accuracy of the coordinate transformations in
\astropysubpkg{coordinates}, we have created a set of benchmarks that we use
to compare transformations between a set of coordinate frames for a number of
packages\footnote{\url{http://www.astropy.org/coordinates-benchmark/summary.html}}.
Since no package can be guaranteed to implement all transformations to
arbitrary precision and some transformations are sometimes subject to
interpretation of standards (in particular in the case of Galactic coordinates),
we do not designate any of the existing packages as the ``ground truth'' but
instead compare each tool to all other tools. The benchmarks are thus useful
beyond the \astropy Project since they allow all of the tools to be compared to
all other tools. The tools included in the benchmark at the moment include the
\astropypkg core package, Kapteyn \citep{kapteyn}, NOVAS \citep{novas},
PALpy \citep{pal}, PyAST \citep[a wrapper for AST, described in][]{ast},
PyTPM\footnote{\url{https://github.com/phn/pytpm}}, PyEphem \citep{pyephem},
and pySLALIB \citep[a Python wrapper for SLALIB, described in][]{slalib}.

@@

The \astropysubpkg{table} subpackage provides functionality for
representing and manipulating heterogeneous data. In some respects,
this is similar to \package{numpy} record arrays \citep{numpy} or
\package{pandas} \texttt{DataFrame} objects \citep{pandas}, but with modifications for
astronomical data. Most notably, tables from \astropysubpkg{table}
allow for table or column metadata and can handle
vectors/arrays or arbitrary objects (with suitable column-like characteristics)
as table entries. The subpackage was described in
detail in \cite{astropy}.  Thus, in what follows, we only summarize
key new features or updates to \astropysubpkg{table} since the
previous \astropy paper. These are support for grouped table
operations, table concatenation, and using array-valued
\package{astropy} objects as table columns.
